Modelling BCG vaccination in the UK

### 2019-12-06 rmds/08-model-development.Rmd
updated TB epi with corrections
model development minor corrections

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the impact of the change in BCG policy on the subset of the population who were directly impacted. Unfortunately, the time horizon of this estimate was limited by the available data. Additionallly, if there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have in-direct impacts, via on-wards transmission, not captured in this estimate. Both of these limitations can be over come using a dynamic transmission model (see Chapter \@ref(introduction)), this explicitly models the rate that individuals are infected using the mass action assumption.[@Anderson1991; @Keeling2007] A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes, via model simulation.

@@

1. **Pulmonary/Extra-Pulmonary TB** - active TB disease can be defined as any symptomatic TB infection but it may present with a range of diverse individual states. Commonly, TB cases are stratified into pulmonary and extra-pulmonary TB cases, with pulmonary cases being individuals who present with TB present in the lungs, and extra-pulmonary TB cases being cases that present with TB symptoms that do not involve the lungs (Chapter \@ref(background)). Often pulmonary cases also present with extra-pulmonary symptoms. It is thought that pulmonary TB cases make up for the vast majority of TB transmission,[@Mathema2018; @Sepkowitz1996] as TB is primarily spread by aerosol transmission, but that extra-pulmonary cases have worse outcomes. The proportion of pulmonary to extra-pulmonary cases has increased over time from 26.2% (1944/7410) in 1982 to 45.8% (2634/5748) in 2016. This may be attributed to the age distribution of TB cases changing, as different age-groups are more likely to progress to pulmonary vs extra-pulmonary TB.[@Lefebvre2017] It may also be related to the increase of non-UK born cases, as a higher proportion of non-UK born cases have extra-pulmonary disease only (51.4%, 2,103/4,089, in 2016), compared to UK born cases (31.9%, 467/1,465, in 2016).[@PHE2017] The model presented here includes both pulmonary and extra-pulmonary cases, with only pulmonary cases contributing to onwards transmission. Extra-pulmonary cases are included so that the full impacts of any intervention can be correctly estimated.

@@

```{r model-flow-diag, out.width = "400px", fig.scap = "Flow diagram for the dynamic TB disease model with demographics and vaccination described.", fig.cap = "Flow diagram for the dynamic TB disease model with demographics and vaccination described. The TB model contains the following compartments; Susceptible ($S$), high risk latent ($H$), low risk latent ($L$), active cases with pulmonary TB ($P$), active TB cases with extra-pulmonary TB only ($E$), pulmonary cases on treatment ($T_P$), and extra-pulmonary cases on treatment ($T_E$). The vaccinated ($v$) and unvaccinated ($u$) populations are represented by $k$, such that $k = u,v$. Age stratification is represented by $a$ (where $a = 1, 2, ...,11$) in the disease model and the $0,\ 1,\ 2,\ 3$ subscripts in the demographic model. Five year age groups are modelled (i.e $0-4$, $5-9$, $10-14$, ...) up to 49 years old, with a single age group for those aged 50-69 years old and those aged 70-89 years old. Individuals aged $90+$ are not explicitly modelled. In the demographic and vaccination model the A compartment represents the demographic processes modelled in all population compartments except for the vaccinated and unvaccinated susceptible populations. Solid arrows represent transition rates within the modelled populations and dashed arrows represent transition rates into, or out of the modelled populations (i.e birth and death processes).", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "model_diagrams", "tb_model_flow.png"))
```

@@

\begin{equation}
\frac{dC}{dt} =  \frac{dC^{d}}{dt} + \frac{dC^{p}}{dt} \\
  (\#eq:split-models)
\end{equation}

The disease model ($C^d$) is then defined as,

@@

Where the unvaccinated ($u$) and vaccinated ($v$) populations are represented by $k = u,v$ and age groups are represented by $a=0,1,2,3, ... 11$. The disease model parameters are defined as follows: $\lambda_a$ is the force of infection; $\epsilon_a$ is the rate of activation from each latent population; $\kappa_a$ is the rate of transition into the low risk latent population; $\nu_a$ is the rate of starting treatment; $\delta$ is the protection from re-infection conferred by prior latent infection; $\Upsilon_a$ is the proportion of cases that develop pulmonary TB, with or without extra-pulmonary TB; $\mu^{P,E}_a$ is the mortality from active pulmonary ($P$) and extra-pulmonary ($E$) TB; $\zeta_a$ is the rate of treatment failure; $\phi_a$ is the rate of successful treatment; $\alpha_a$ is the effectiveness of the BCG vaccine at preventing active TB disease; and $\chi^k_a$ is the protection inferred due to vaccination from initial infection. In the unvaccinated population $\alpha = 0$ and $\chi = 0$. Parameters with an $a$ subscript, or superscript, are age-stratified.

The demographic and vaccination model ($C^p$) is then defined as ($A = H,\ L,\ P,\ E,\ T_P,\ T_E$),

@@

\begin{equation}
\frac{dS^{vp}_{a}}{dt} = (1 - sgn(a)) \gamma_a\omega(t) + sgn(a)\gamma_a\theta_{a - 1} S^u_{a - 1} - \theta_a S^v_a -\mu_{a}(t) S^v_a
  (\#eq:sus-age-v-model)
\end{equation}

\begin{equation}
\frac{dA^{kp}_{a}}{dt} = sgn(a)\theta_{a - 1} A^k_{a - 1} - \theta_a A^k_{a} - \mu_{a}(t) A^k_{a}
  (\#eq:other-age-model)
\end{equation}

@@

The signum function used above is defined as follows;
\begin{equation}
sgn(x) := \begin{cases}
0 & \text{if } x = 0, \\
1 & \text{if } x > 0. \end{cases}
  (\#eq:signum)
\end{equation}

@@

\begin{equation}
\lambda^k_a = \frac{\beta_a}{N} \sum\limits_{i=1}^{A} \rho_i C_{ai}\left(\frac{M_{a}\iota_i}{\nu_a^P} + \sum\limits_{j = u,v}P^j_i\right)
  (\#eq:force-of-infection)
\end{equation}

Where $\iota_a$ is the age stratified number of non-UK born pulmonary cases notified in a given year, $P^{u,v}_a$ is the number of vaccinated, and unvaccinated, pulmonary TB cases, $\rho_a$ is the age-specific proportion of cases that are smear positive, $\nu^P_a$ is the age-specific rate of starting treatment for active pulmonary TB, $C_{ai}$ is the age-stratified contact matrix (Section \@ref(contact-matrix)), $\beta_a$ is the age-stratified transmission probability, and $M_a$ is the age-stratified mixing rate between the UK born and non-UK born population.

@@

The effectiveness of the BCG vaccine is usually estimated using its effectiveness at reducing the incidence of active TB cases in a susceptible population. In the model outlined in this chapter the action of the BCG vaccine has been split into its main effect of reducing the rate of latent TB cases developing active disease and its secondary effect of reducing the likelihood of initial infection. There are few estimates of the effectiveness of the BCG vaccine at preventing active TB in cases that are already latently infected and where these estimates do exist they are not stratified by time since vaccination, or age at vaccination.[@Roy2014] The overall effectiveness ($\alpha^T_a$) of the BCG vaccine can be estimated from the combined effectiveness at preventing initial infection ($\chi_a$) and then preventing activation in latently infected individuals ($\alpha_a$) using the following equation,

\begin{equation}
  \alpha^T_a = \chi_a + (1 - \chi_a) \alpha_a
  (\#eq:full-bcg-effectiveness)
\end{equation}

@@

\begin{equation}
  \alpha_a = \frac{\alpha^T_a - \chi_a}{1 - \chi_a}
  (\#eq:latent-active-bcg-effectiveness)
\end{equation}

There is strong evidence that the overall effectiveness of the BCG vaccine reduces over time.[@Abubakar2013; @Mangtani2017] For this reason the effectiveness of the BCG vaccination overall ($\alpha^T_a$) has been stratified by the time since vaccination (by 5 year age groups). This step-wise approach has been chosen as the majority of studies report estimates for these groups and the precise functional form of the reduction in protection is unknown. For 0-4, and 5-9, years since vaccination estimates of the effectiveness of the BCG vaccine were extracted from the MRC trial.[@Hart1972] Using the published data, Poisson regression was used to estimate Rate Ratios and 95% confidence intervals. For 10-29 years after vaccination Rate Ratio estimates from a more recent case control cohort study in the UK born vaccinated at school-age have been used.[@Mangtani2017] Table \@ref(tab:bcg-eff-est) details the estimated effectiveness for each five yearly band after initial vaccination. I have assumed that the BCG vaccine is equally effective regardless of when the vaccine is given as there is no evidence that protection reduces when given to older age groups in England.

```{r bcg-eff-est}
kable(eff_bcg, booktabs = TRUE, escape = TRUE,
      caption.short = "Estimates of the effectiveness of the BCG vaccine at preventing active TB disease stratified by years since vaccination.", 
      caption = "Estimates of the effectiveness of the BCG vaccine at preventing active TB disease stratified by years since vaccination. For 0-9 years since vaccination estimates were derived using Poisson regression from the MRC BCG trial and for 10-29 years since vaccination estimates were extracted from a more recent case control cohort study in the UK born vaccinated at school-age. BCG effectiveness at preventing active TB disase used as prior distributions in the model.") %>% 
  kable_styling(latex_options = "hold_position")
```

Using the literature derived estimates for the Risk Ratio (RR) of the BCG vaccine at different periods after vaccination, the log normal approximation for the distribution of Risk Ratios, and the relationship between vaccination effectiveness and the Risk Ratio (Effectiveness = 1 - Risk Ratio) I derived a prior distribution for the overall effectiveness of the BCG vaccine ($\alpha^T_a$). This can be summarised by the following equation,

\begin{equation}
  \alpha^T_a = 1 - \text{exp}\left(\mathcal{N}(\text{ln}(RR_{t_v}, SE_{t_v})\right)
  (\#eq:rr-to-alpha)
\end{equation}

@@

\begin{equation}
  \chi_j = \frac{\alpha^T_j\chi_i}{\alpha^T_i}
  (\#eq:fun-prev-initial-inf)
\end{equation}

@@

```{r mortality-england, fig.scap = "3 year rolling average expected remaining lifespan stratified by age group in England from 2000 to 2014.", fig.cap = "3 year rolling average expected remaining lifespan stratified by age group in England from 2000 to 2014. Age is grouped into 5 year age groups from 0 to 49, from 50-69, and from 70 to 89. Those aged 90+ are excluded due to low quality data. The age groups used here represent those used in the model. Data from this figure was sourced from the ONS age-specific mortality rate estimates with projections based on an age-stratified exponential model.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "mortality.png"))
```

@@

The semi-stochastic transmission dynamic model of TB transmission and BCG vaccination presented in this chapter provides a detailed overview of the features required to reproduce the observed epidemiology of TB in England. The model was robustly parameterised using routine surveillance data where available and otherwise using literature sources. The assumptions required by the model can be explored by fitting the model to observed data and assessing the goodness of fit. This is the focus of the next chapter. In addition the model may also be used to explore the impact of current and historic BCG vaccination policy, both in the observed data and projected into the future. This is explored in the final chapter of this thesis.

>>>>

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the impact of the change in BCG policy on the subset of the population who were directly impacted. Unfortunately, the time horizon of this estimate was limited by the available data. Additionallly, if there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have indirect impacts, via onwards transmission, not captured in this estimate. Both of these limitations can be overcome using a dynamic transmission model (see Chapter \@ref(introduction)), this explicitly models the rate that individuals are infected using the mass action assumption.[@Anderson1991; @Keeling2007] A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes, via model simulation.

@@

1. **Pulmonary/Extra-Pulmonary TB** - active TB disease can be defined as any symptomatic TB infection but it may present with a range of diverse individual states. Commonly, TB cases are stratified into pulmonary and extra-pulmonary TB cases, with pulmonary cases being individuals who present with TB present in the lungs, and extra-pulmonary TB cases being cases that present with TB symptoms that do not involve the lungs (Chapter \@ref(background)). Often pulmonary cases also present with extra-pulmonary symptoms. It is thought that pulmonary TB cases make up for the vast majority of TB transmission,[@Mathema2018; @Sepkowitz1996] as TB is primarily spread by aerosol transmission, but that extra-pulmonary cases have worse outcomes. The proportion of pulmonary to extra-pulmonary cases has increased over time from 26.2% (1944/7410) in 1982 to 45.8% (2634/5748) in 2016. This may be attributed to the age distribution of TB cases changing, as different age-groups are more likely to progress to pulmonary TB than other age-groups.[@Lefebvre2017] It may also be related to the increase of non-UK born cases, as a higher proportion of non-UK born cases have extra-pulmonary disease only (51.4%, 2,103/4,089, in 2016), compared to UK born cases (31.9%, 467/1,465, in 2016).[@PHE2017] The model presented here includes both pulmonary and extra-pulmonary cases, with only pulmonary cases contributing to onwards transmission. Extra-pulmonary cases are included so that the full impacts of any intervention can be correctly estimated.

@@

```{r model-flow-diag, out.width = "400px", fig.scap = "Flow diagram for the dynamic TB disease model with demographics and vaccination described.", fig.cap = "Flow diagram for the dynamic TB disease model with demographics and vaccination described. The TB model contains the following compartments; Susceptible ($S$), high risk latent ($H$), low risk latent ($L$), active cases with pulmonary TB ($P$), active TB cases with extra-pulmonary TB only ($E$), pulmonary cases on treatment ($T_P$), and extra-pulmonary cases on treatment ($T_E$). The vaccinated ($v$) and unvaccinated ($u$) populations are represented by $k$, such that $k = u,v$. Age stratification is represented by $a$ (where $a = 1, 2, ...,11$) in the disease model and the $0,\ 1,\ 2,\ 3$ subscripts in the demographic model. Each age groups spans 5 years (i.e $0-4$, $5-9$, $10-14$, ...) up to 49 years old, with a single age group for those aged 50-69 years old and those aged 70-89 years old. Individuals aged $90+$ are not explicitly modelled. In the demographic and vaccination model the A compartment represents the demographic processes modelled in all population compartments except for the vaccinated and unvaccinated susceptible populations. Solid arrows represent transition rates within the modelled populations and dashed arrows represent transition rates into, or out of the modelled populations (i.e birth and death processes).", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "model_diagrams", "tb_model_flow.png"))
```

@@

\begin{equation}
\frac{dC}{dt} =  J_{d}C + J_{p}C
  (\#eq:split-models)
\end{equation}

The disease model ($J_{d}C$) is then defined as,

@@

Where the unvaccinated ($u$) and vaccinated ($v$) populations are represented by $k = u,v$ and age groups are represented by $a=0,1,2,3, ... 11$. The disease model parameters are defined as follows: $\lambda_a$ is the force of infection; $\epsilon_a$ is the rate of activation from each latent population; $\kappa_a$ is the rate of transition into the low risk latent population; $\nu_a$ is the rate of starting treatment; $\delta$ is the protection from re-infection conferred by prior latent infection; $\Upsilon_a$ is the proportion of cases that develop pulmonary TB, with or without extra-pulmonary TB; $\mu^{P,E}_a$ is the mortality from active pulmonary ($P$) and extra-pulmonary ($E$) TB; $\zeta_a$ is the rate of treatment failure; $\phi_a$ is the rate of successful treatment; $\alpha_a$ is the effectiveness of the BCG vaccine at preventing active TB disease; and $\chi^k_a$ is the protection inferred due to vaccination from initial infection. In the unvaccinated population (when $k=u$) $\alpha^u_a = 0$ and $\chi^u_a = 0$. Parameters with an $a$ subscript, or superscript, are age-stratified.

The demographic and vaccination model ($J_{p}C$) is then defined as ($A = H,\ L,\ P,\ E,\ T_P,\ T_E$),

@@

\begin{equation}
\frac{dS^{vp}_{a}}{dt} = (1 - \text{sgn}(a)) \gamma_a\omega(t) + \text{sgn}(a)\gamma_a\theta_{a - 1} S^u_{a - 1} - \theta_a S^v_a -\mu_{a}(t) S^v_a
  (\#eq:sus-age-v-model)
\end{equation}

\begin{equation}
\frac{dA^{kp}_{a}}{dt} = \text{sgn}(a)\theta_{a - 1} A^k_{a - 1} - \theta_a A^k_{a} - \mu_{a}(t) A^k_{a}
  (\#eq:other-age-model)
\end{equation}

@@

The signum function used above is defined as follows;
\begin{equation}
\text{sgn}(x) := \begin{cases}
0 & \text{if } x = 0, \\
1 & \text{if } x > 0. \end{cases}
  (\#eq:signum)
\end{equation}

@@

\begin{equation}
\lambda_a = \frac{\beta_a}{N_a} \sum\limits_{i=1}^{A} \rho_i C_{ai}\left(\frac{M_{i}\iota_i}{\nu_i^P} + \sum\limits_{j = u,v}P^j_i\right)
  (\#eq:force-of-infection)
\end{equation}

Where $\iota_i$ is the age stratified number of non-UK born pulmonary cases notified in a given year, $P^{u,v}_i$ is the number of vaccinated, and unvaccinated, pulmonary TB cases, $\rho_i$ is the age-specific proportion of cases that are smear positive, $\nu^P_i$ is the age-specific rate of starting treatment for active pulmonary TB, $C_{ai}$ is the age-stratified contact matrix (Section \@ref(contact-matrix)), $\beta_a$ is the age-stratified transmission probability, $N_a$ is the number of people in age group $a$, and $M_i$ is the age-stratified mixing rate between the UK born and non-UK born population.

@@

The effectiveness of the BCG vaccine is usually estimated using its effectiveness at reducing the incidence of active TB cases in a susceptible population. In the model outlined in this chapter the action of the BCG vaccine has been split into its main effect of reducing the rate of latent TB cases developing active disease and its secondary effect of reducing the likelihood of initial infection. There are few estimates of the effectiveness of the BCG vaccine at preventing active TB in cases that are already latently infected and where these estimates do exist they are not stratified by time since vaccination, or age at vaccination.[@Roy2014] The overall effectiveness ($\alpha^T_a$) of the BCG vaccine can be estimated from the combined effectiveness at preventing initial infection ($\chi^v_a$) and then preventing activation in latently infected individuals ($\alpha^v_a$) using the following equation,

\begin{equation}
  \alpha^T_a = \chi^v_a + (1 - \chi^v_a) \alpha^v_a
  (\#eq:full-bcg-effectiveness)
\end{equation}

@@

\begin{equation}
  \alpha^v_a = \frac{\alpha^T_a - \chi^v_a}{1 - \chi^v_a}
  (\#eq:latent-active-bcg-effectiveness)
\end{equation}

There is strong evidence that the overall effectiveness of the BCG vaccine reduces over time.[@Abubakar2013; @Mangtani2017] For this reason the effectiveness of the BCG vaccination overall ($\alpha^T_a$) has been stratified by the time since vaccination (by 5 year age groups). This step-wise approach has been chosen as the majority of studies report estimates for these groups and the precise functional form of the reduction in protection is unknown. For 0-4, and 5-9, years since vaccination estimates of the effectiveness of the BCG vaccine were extracted from the MRC trial.[@Hart1972] Using published data, Poisson regression was used to estimate Rate Ratios and 95% confidence intervals. For 10-29 years after vaccination Rate Ratio estimates from a more recent case control cohort study in the UK born vaccinated at school-age have been used.[@Mangtani2017] Table \@ref(tab:bcg-eff-est) details the estimated effectiveness for each five yearly band after initial vaccination from these data sources. I have assumed that the BCG vaccine is equally effective regardless of when the vaccine is given as there is no evidence that protection reduces when given to older age groups in England.

```{r bcg-eff-est}
kable(eff_bcg, booktabs = TRUE, escape = TRUE,
      caption.short = "Estimates of the effectiveness of the BCG vaccine at preventing active TB disease stratified by years since vaccination.", 
      caption = "Estimates of the effectiveness of the BCG vaccine at preventing active TB disease stratified by years since vaccination. For 0-9 years since vaccination estimates were derived using Poisson regression from the MRC BCG trial and for 10-29 years since vaccination estimates were extracted from a more recent case control cohort study in the UK born vaccinated at school-age.") %>% 
  kable_styling(latex_options = "hold_position")
```

Using the literature derived estimates for the Risk Ratio (RR) of the BCG vaccine at different periods after vaccination, the log normal approximation for the distribution of Risk Ratios, and the relationship between vaccination effectiveness and the Risk Ratio (Effectiveness = 1 - Risk Ratio) I derived a prior distribution - which can be sampled from - for the overall effectiveness of the BCG vaccine ($\alpha^T_a$). This can be summarised by the following equation,

\begin{equation}
  \alpha^T_a = 1 - e^{\left(\mathcal{N}(\text{ln}(RR_{t_v}), SE_{t_v})\right)}
  (\#eq:rr-to-alpha)
\end{equation}

@@

\begin{equation}
  \chi^v_j = \frac{\alpha^T_j\chi^v_i}{\alpha^T_i}
  (\#eq:fun-prev-initial-inf)
\end{equation}

@@

```{r mortality-england, fig.scap = "Three year rolling average expected remaining lifespan stratified by age group in England from 2000 to 2014.", fig.cap = "Three year rolling average expected remaining lifespan stratified by age group in England from 2000 to 2014. Age is grouped into 5 year age groups from 0 to 49, from 50-69, and from 70 to 89. Those aged 90+ are excluded due to low quality data. The age groups used here represent those used in the model. Data from this figure was sourced from the ONS age-specific mortality rate estimates with projections based on an age-stratified exponential model.", out.extra = ""}
knitr::include_graphics(file.path(resources_path, "resources", "figure", "mortality.png"))
```

@@

The semi-stochastic transmission dynamic model of TB transmission and BCG vaccination presented in this chapter provides a detailed overview of the features required to reproduce the observed epidemiology of TB in England. The model was robustly parameterised using routine surveillance data where available and otherwise using literature sources. The assumptions required by the model can be explored by fitting the model to observed data and assessing the goodness of fit. This is the focus of the next chapter. In addition the model may also be used to explore the impact of current and historic BCG vaccination policy, both in the observed data and projected into the future.Both of these scenarios are explored in the final chapter of this thesis.
